How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?

### 2024-02-12 outputs/paper/paper.qmd
updated paper

>>>>

The COVID-19 pandemic starkly highlighted how social vulnerabilities and disparities in healthcare access exacerbated the impact of global health crises, contributing to significant variations in disease outcomes and underscoring the need for targeted public health strategies that address these disparities' root causes. Social vulnerabilities, such as socioeconomic status, race, ethnicity, and housing conditions, critically determined COVID-19 outcomes' severity, with populations in crowded housing, limited access to sanitation, and lower socioeconomic brackets experiencing higher transmission rates due to social distancing and hygiene maintenance challenges. An independent study done by the `Government of Canada` states that it "identified that the risk of COVID-19 related deaths in Black, Asian and minority ethnic groups was nearly 1.5 times higher than White individuals" [@citesource5]. Another news article done by the `MSN` [@citesource6] states that 'racial minority participants reported more negative impacts on health status, activity, and absence from work as compared to the White population.' The pandemic's economic toll further limited these groups' healthcare access, amplifying vulnerabilities. Disparities in healthcare access played a significant role in influencing COVID-19 morbidity and mortality, with communities facing healthcare facility shortages, provider scarcities, and barriers due to insurance or financial constraints at heightened risk. These disparities were evident in the uneven vaccine distribution and access, highlighting the advantages of regions with strong healthcare infrastructure. Marginalized populations, including racial and ethnic minorities, faced compounded risks from social vulnerabilities and healthcare disparities, evidenced by higher infection, hospitalization, and death rates due to factors like essential service employment and prevalent pre-existing conditions. A news article from `CNN` [@citesource7] talks about how many essential workers, who cannot work from home, are from black and Latinx communities. These include healthcare professionals, grocery cashiers, delivery workers, and public transport employees. Despite their crucial roles, they often lack adequate pay, protection, and respect. Addressing these disparities in future pandemics requires public health strategies that prioritize equity and inclusivity, including community-based healthcare investments, enhanced vulnerable community outreach, and equitable healthcare resource access policies. By incorporating social determinants of health into public health preparedness plans, responses can effectively protect at-risk populations, making future public health responses more resilient, inclusive, and effective in safeguarding all population segments.
